---
title: "R HW 02: Data Visualization"
output: 
  tufte::tufte_html:
    css: "lab.css"
    tufte_variant: "envisioned"
    highlight: pygments
    toc: true
    toc_depth: 1
link-citations: yes
editor_options: 
  markdown: 
    wrap: sentence
---

## Goals

In this lab, you will...

- Gain proficiency in data visualization
- Apply principles of effective data visualization to a real dataset
- Continue developing a workflow for reproducible data analysis.

## Getting started 

-   Download R_HW_02.qmd and anes_2020_subset.csv from our course website (under Assignments)
-   Place the downloaded .qmd document in your STAT_7500 folder on your computer
-   Place the .csv file inside a subfolder called 'data' inside the STAT_7500 folder
-   From inside your STAT_7500 RStudio Project, open and render the Quarto document

# Packages

We will work with the `tidyverse` package as usual. We will also use `viridis` and the `ggridges` packages.

```{r load-packages, message = FALSE, echo = FALSE, warning = FALSE}
library(tidyverse)
library(viridis)
library(ggridges)
```

# Data: 2020 American National Election Study

The data for this homework assignment comes from the [2020 American National Election Study](https://electionstudies.org/data-center/2020-time-series-study/). Note, you will need the .csv file to be located inside a subfolder called data for this code to work properly. 

```{r, eval = FALSE}
anes <- read_csv("data/anes2020_subset.csv")
```

A subset of variables are provided here. Some of them have been recoded, while others you may need to recode in order to be able to carry out your analysis. The variables are as follows:

- `CASEID`: a Case ID for the respondent.

- `hunt_fish`: a dummy variable asking if the respondent has gone hunting or fishing in the past year.

- `scientists`: A feeling thermometer question that asks how warmly respondents feel towards scientists. A score of 0 represents the coolest rating, while a score of 100 represents the warmest rating.

- `education`: An ordinal categorical representing the highest level of education for the respondent, ranging from less than high school to a professional degree.

- `ideology`: a seven point self-rating scale for the respondent’s ideology ranging from most liberal to most conservative

- `urbanrural`: a variable indicating how rural or urban the respondent’s home community is with four possible values: rural, small town, suburb, or city.

# Exercises 

All plots should follow the best visualization practices discussed in lecture. Plots should include an informative title, axes should be labeled, and careful consideration should be given to aesthetic choices.

Remember that continuing to develop a sound workflow for reproducible data analysis is important as you complete the lab and other assignments in this course. You should provide adequate comments to your code and narrative text interpreting your results.  

1. Provide a brief description of the data collection process - who collected the data, when was it collected, who was surveyed, and anything else you find relevant/interesting. 

2. How many rows are in the `anes` dataset? How many columns? Please include code and output to support your response. 

3. Create a bar chart showing the `ideology` of the respondents, with the count on the y-axis. 
  - Please remember to include labels. 
  - What is the most common ideology? 
  - Do respondents tend to be moderate or more ideologically extreme?

4. Now, let’s examine whether ideologies are different based upon where people live. Please make a filled bar plot, showing one bar for each `ideology`, with the percentage of respondents on the y-axis going from 0-1, and the fill determined by `urbanrural`. 
  - You are encouraged but not required to use `viridis` colors. 
  - Please remember to include labels. 
  - Where do people of different ideologies tend to live? 
  - Does the percentage of non-responses (i.e., people who said NA) vary much by ideology?
  
5. How do people view scientists? Please make a histogram with the feeling thermometer on the x-axis and the number of respondents on the y-axis. Then, please comment on features of the histogram such as skewness and peaks.

6. Does the ideology of those who have gone hunting or fishing in the past year differ from those who haven’t? Please make side-by-side boxplots of these two groups.
  - You should start your code with:

```{r, eval = FALSE}
anes |>
  drop_na(hunt_fish) |>
  mutate(hunted_fished = ifelse(hunt_fish == 0, "Did Not Hunt or Fish",  
                                "Hunted or Fished")) |>
```

```{marginfigure}
You can read more about ridge plots [here](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html).
```

7. Explore the same question as in Exercise 6, this time using `geom_density_ridges()` to construct a ridge plot as an alternative to side-by-side box plots.  

8. Describe what you observe in both plots from Exercises 6 & 7 and what you learn from one plot that you do not see in the other or that adds additional context to the other.

9. Is there a relationship between a respondent’s education level and how they view scientists? Please make a scatterplot of these two variables with education on the x-axis and view of scientists on the y-axis. 
  - Include a line of best fit using a `geom_smooth()` layer with the option `method = "lm"`. 
  - Do you think this is an especially useful visualization? Why or why not?

10. What is an additional question you could investigate using this data? State the question, provide a visualization that investigates it, and comment on what your visualization shows. 

11. Write a brief paragraph summarizing what you found in this lab, what limitations might exist to the data/analyses, and what additional data you would like to have to explore further questions. 

# Submission

Before submitting your assignment on Blackboard, Render one last time, then open up your .html in a full window and inspect the following:

+ Check your code for neatness - add spaces and line breaks where appropriate to improve readability

+ Check any visualizations for informative and clean labels and titles

+ Check any tables for clean formatting (e.g. with `kbl() |> kable_minmal()`)

+ Check that you've interpreted output as appropriate

+ Make sure extraneous warnings and messages are suppressed (e.g. set `#| warning: false`, `#| message: false`)

+ Check that Exercises are clearly labeled and your text responses are visually distinguished

+ Check for overall neatness & organization of the report (e.g. headers, subheaders, use of bullets, font changes)

To submit, export your html as a .pdf and upload the pdf document to the relevant assignment on Blackboard. Refer back to Lab 01 for detailed instructions on how to submit your assignment. 

# Grading (10 pts)

<br>

| Component | Points |
|:----------|:-------|
| Ex 1      | 0.5      |
| Ex 2      | 0.5    |
| Ex 3      | 0.5    |
| Ex 4      | 1     |
| Ex 5      | 1      |
| Ex 6      | 1    |
| Ex 7      | 1    |
| Ex 8      | 1    |
| Ex 9      | 1    |
| Ex 10      | 1    |
| Ex 11     | 1    |
| Workflow & formatting | 0.5     |

**Grading notes**:

-   The "Workflow & formatting" grade is to assess the reproducible workflow. This includes having readable code (e.g. adequate use of spacing and line breaks), labeled code chunks, informative headers and sub-headers, and an overall organized and uncluttered report (e.g. suppress messages & warnings, no extraneous output). 

