---
title: "R HW 02 - Working with Data & Data Visualization"
format:
  html:
    css: styles.css
    grid:
      margin-width: 350px
fig-numbering: false
toc: true
toc-location: right
categories:
  - Data importing
  - Data inspection
  - Data visualization
  - ggplot2
editor_options:
  chunk_output_type: console
callout-icon: false
---

# Goals

The main goal of this assignment is to practice a reproducible workflow for working with data in R: importing data, checking that R interpreted it appropriately, analyzing relevant variables, and communicating what you learn.

You will...

+ Import data from an external file using a relative path
+ Inspect the structure and contents of imported data
+ Use data documentation to identify values that represent missingness
+ Distinguish between how R stores a variable and what the variable means
+ Use vectors, indexing, and summary statistics to investigate change over time
+ Export and verify a prepared data file
+ Create and interpret scatterplots
+ Map variables to aesthetic features of a visualization
+ Compare color and faceting as alternative ways to display a third variable

The individual R commands matter, but the larger goal is to develop a reliable process: understand where the data came from, inspect before analyzing, make purposeful choices, and interpret results in context.

# Getting Started

To get started on this assignment, do the following:

- Open our class [GitHub repo](https://github.com/kgfitzgerald/STAT_7500_templates)
- Navigate to `R_HW_02.qmd`
- Click the download icon (hover text should say "Download raw file")
- Place the downloaded `.qmd` document in your STAT_7500 folder
- Download `patients.csv` and place it inside the `data` folder in your STAT_7500 project
- From inside your STAT_7500 RStudio Project, open and render the Quarto document

Your project should be organized approximately like this:

```         
STAT_7500/
├── R_HW_02.qmd
└── data/
    └── patients.csv
```

The document you are currently reading contains all the instructions for R HW 02. You will complete the assignment by adding code and text responses to the `R_HW_02.qmd` document in RStudio.

::: column-margin :::
Keeping the data and analysis files together in an R Project allows you to use a relative file path such as `"data/patients.csv"`. Do not use a file path that begins with a user name or refers to a location specific to your computer.
:::

## Packages

In this assignment, we will use **tidyverse** for importing and visualizing data and **palmerpenguins** for the `penguins` dataset.

You can run `install.packages("nameofpackage")` in your Console if you do not already have a package installed. Then, load the packages in your Quarto document:

```{r message=FALSE}
library(tidyverse)
library(palmerpenguins)
```

# Data: Patient Health-Awareness Scores

The patient data were assembled from instructional files previously used in the SAS portion of STAT 7500. They should be treated as instructional data rather than as records from a documented research study.

The file contains records for thirty patients who participated in a six-week health-awareness program. Health-awareness surveys were administered before and after the program. Scores range from 0 to 20, with higher scores indicating greater awareness of healthy habits.

The variables in `patients.csv` are:

- `patient_id`: unique patient identification code
- `gender_code`: gender recorded as `0` for male and `1` for female
- `baseline_score`: health-awareness score before the program
- `followup_score`: health-awareness score after the program

The data documentation also tells us that:

- `Not recorded` represents a missing gender value; and
- `.` represents a missing survey score.

# Exercises

## Importing and inspecting data

::: callout-note
### Exercise 1

Import `patients.csv` using `read_csv()` and assign the resulting data frame to an object named `patients_raw`.

Use a relative file path. Then, use `glimpse()` to inspect the imported data.
:::


::: column-margin :::
After importing data, always inspect the result. The absence of an error message does not guarantee that R interpreted every variable as intended.
:::

::: callout-note
### Exercise 2

+ How many observations and variables are in the data? Use code to support you answer. 
+ What does one row represent?
+ Which variables did R interpret as character variables even though their nonmissing values are numerical codes or scores? Why did this occur?
+ Why is it important to answer these questions before beginning an analysis?
:::

`r emo::ji("pause_button")` `r emo::ji("right_arrow")` ✅ *This is a good place to pause and Render to make sure there are no errors in your .qmd*

## Identifying missing values during import

Missing data in R are represented by `NA`. However, external files may use many different values to indicate missingness. We should use the data documentation—not guesswork—to determine what those values mean.

::: callout-note
### Exercise 3

Import `patients.csv` again and assign the result to an object named `patients`.

This time, use the `na` argument in `read_csv()` to identify `"Not recorded"` and `"."` as missing values.

Use `glimpse()` and `summary()` to inspect `patients`, and then respond to the following:

+ Which variables changed type and why?
+ Did identifying the missing values change the number of rows or columns? Why or why not?
+ Although `gender_code` is now stored as a numerical variable, should it be interpreted as a quantitative measurement? Explain.
+ What does this example demonstrate about the difference between a variable's storage type and its meaning?
:::

```{r}

```

::: column-margin :::
A value such as `.` is not universally a missing value. We treat it as missing here because the data documentation tells us what it represents in this particular file.
:::

`r emo::ji("pause_button")` `r emo::ji("right_arrow")` ✅ *This is a good place to pause and Render to make sure there are no errors in your .qmd*

## Investigating change over time

The survey was administered to the same patients before and after the program. Therefore, each patient's follow-up score should be compared with that same patient's baseline score.

::: callout-note
### Exercise 4

Create a vector called `score_change` by subtracting each patient's baseline score from their follow-up score:

```r
score_change <- patients$followup_score - patients$baseline_score
```

Then:

+ Print `score_change`.
+ Print only the fifth element using `[]` notation.
+ Verify that `score_change` is numerical using `class()`.
+ Use `summary()` to inspect the distribution of the change scores.
+ Use appropriate summary statistic function (e.g. `mean()`) to reproduce the mean, median, minimum, and maximum printed out by `summary()`. Handle missing values appropriately.
+ Explain why `score_change` contains a missing value. Does that missing value represent a patient whose score did not change?
+ In 2–3 sentences, describe what the results suggest about changes in health-awareness scores.
+ Do these summaries alone establish that the program *caused* the observed changes? Explain briefly.
:::


::: column-margin :::
A positive value of `score_change` indicates that the patient's score increased. A negative value indicates that it decreased. A value of zero indicates no change.
:::

## Exporting and verifying data

::: callout-note
### Exercise 5

Use `write_csv()` to export `patients` as `data/patients_clean.csv`.

Then:

+ Confirm that `patients_clean.csv` appears inside the `data` folder in your project.
+ Import the exported file and assign it to an object named `patients_check`.
+ Use `dim()` to verify that `patients` and `patients_check` have the same dimensions.
+ Use `glimpse()` to verify that the variables were imported with the expected types.
+ Explain why checking an exported file is a useful part of a reproducible workflow.
:::

`r emo::ji("pause_button")` `r emo::ji("right_arrow")` ✅ *This is a good place to pause and Render to make sure there are no errors in your .qmd*

# Data Visualization

For the remaining exercises, use the `penguins` data from the **palmerpenguins** package. Each row represents an individual penguin observed in the Palmer Archipelago in Antarctica.

Before constructing a visualization, use `glimpse()` to remind yourself of the dataset's structure.

```{r}
glimpse(penguins)
```

::: callout-note
### Exercise 6

**What is the relationship between flipper length and body mass in penguins from the Palmer Archipelago in Antartica?**

Construct a scatterplot with:

+ `flipper_length_mm` on the x-axis; and
+ `body_mass_g` on the y-axis.

Include an informative title and descriptive axis labels with the units of measurement.

Finally, respond to the following:

+ What relationship do you observe between flipper length and body mass?
+ Did R produce a warning? If so, what does it mean?
:::

```{r fig.fullwidth=TRUE}

```

::: column-margin :::
Recall the basic structure:

```r
ggplot(
  data = data_frame,
  aes(x = x_variable,
      y = y_variable)
) +
  geom_point()
```
:::

## Mapping an aesthetic

::: callout-note
### Exercise 7

Modify your plot from Exercise 6 so that dot color represents penguin species.

Then, respond to the following:

+ What additional information does color reveal?
+ Does the overall relationship appear to be explained partly by differences among species? Explain using evidence from the visualization.
+ In your code, was color **mapped** to a variable or **set** to a constant value? How do you know?
:::

```{r fig.fullwidth=TRUE}

```

## Comparing color and faceting

::: callout-note
### Exercise 8

Create another version of the scatterplot that uses separate facets for the three penguin species instead of mapping species to color.

Compare the faceted plot with the color-coded plot from Exercise 7.

+ What is easier to see in the color-coded plot?
+ What is easier to see in the faceted plot?
+ Which version would you use to communicate the relationship among species, flipper length, and body mass? Briefly justify your choice based on your communication goal.
:::

```{r fig.fullwidth=TRUE}

```

`r emo::ji("pause_button")` `r emo::ji("right_arrow")` ✅ *This is a good place to pause and Render to make sure there are no errors in your .qmd*

## Mapping versus setting

Consider the following code:

```r
ggplot(
  penguins,
  aes(
    x = flipper_length_mm,
    y = body_mass_g,
    color = "navyblue"
  )
) +
  geom_point()
```

::: callout-note
### Exercise 9

Before running the code above, predict what it will produce. Then, copy it into an executable code chunk, run it, and compare the result with your prediction.

+ Why does the plot contain a legend?
+ Why are the points not necessarily navy blue?

Correct the code so that all points are navy blue and no color legend is created. Then, modify the corrected plot so that all points:

+ are navy blue;
+ are slightly transparent; and
+ have a size larger than the default.

Explain why `color`, `alpha`, and `size` belong inside or outside `aes()` in your final plot. Your explanation should describe the general rule, not only this particular example.
:::



`r emo::ji("right_arrow")` ✅ ⬆️ *Yay, you're done! Render one last time and verify that everything is displayed correctly in your R_HW_02.html file. To view the .html, click on the file name in the Files quadrant and select "View in Web Browser". If everything looks good, you're ready to submit!*

# Submit

::: {.callout-important}

Unfortunately, Brightspace does not handle .html files well. Before submitting, you will need to convert your .html file to a .zip folder. Locate the file on your computer (not in RStudio), right click, and select "Compress." Then, upload the .zip folder to Brightspace.

:::

# Grading (10pts)

Homework assignments are opportunities to practice the skills and habits you will use on larger course projects. They are graded holistically rather than point-by-point.

Your homework will be evaluated for:

* **Completion & substantive correctness (8 points):** You have attempted all exercises, your code and results are generally correct, and your written responses appropriately interpret and engage with the data.
* **Reproducible workflow & presentation (2 points):** Your document renders successfully and your code, output, visualizations, and responses are organized and readable.

Minor mistakes will not be penalized individually. You may also receive feedback about misconceptions or habits to improve even when they do not result in a point deduction. Use this feedback in future homework assignments and projects.

### Homework rubric - 10 points total

#### Completion & substantive correctness — 8 points

| Score | Description |
|---|---|
| **8** | Complete and substantively correct. Code and analyses accomplish what was asked, and written responses appropriately interpret the results. Minor imperfections are okay. |
| **7** | Mostly complete and correct, with one notable error, omission, or misunderstanding. |
| **6** | Several errors or omissions, but the work demonstrates reasonable understanding of the main ideas. |
| **4–5** | Significant errors, omissions, or misunderstandings that should be addressed before moving on. |
| **0–3** | Substantially incomplete, incorrect, or shows limited engagement with the homework. |

#### Reproducible workflow & presentation — 2 points

| Score | Description |
|---|---|
| **2** | Document renders successfully; code, output, visualizations, and written responses are reasonably organized and readable. |
| **1** | Some problems with reproducibility, organization, readability, or presentation. |
| **0** | Major problems with reproducibility, organization, readability, or presentation. |

**Total: 10 points**
