---
title: "R HW 05 - National Women's Soccer"
format:
  html:
    css: styles.css
    grid:
      margin-width: 350px
fig-numbering: false
toc: true
toc-location: right
categories:
  - Data wrangling
  - Joins
  - Sports analytics
  - National Women's Soccer League
editor_options:
  chunk_output_type: console
callout-icon: false
---

# Goals

In this assignment, you will:

+ Develop proficiency with data-wrangling functions
+ Join multiple datasets from the National Women's Soccer League
+ Use wrangling and visualization to answer questions about team and player performance for the Portland Thorns FC

::: {.column-margin}
Joins allow us to combine information from multiple data sources into a single analytical workflow.
:::

---

# Getting Started

+ Download the `.qmd` file for this assignment from our class GitHub repo

# Packages

We will primarily use the **tidyverse** package.
Additional packages provide color palettes, table formatting,
ridge plots, and access to NWSL data.

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse)
library(viridis)
library(kableExtra)
library(ggridges)
library(devtools)
devtools::install_github("nwslR/nwslR", ref = "v0.0.0.9002")
library(nwslR)
```

# Data: The Portland Thorns FC

For this assignment, you will work with data about the
[Portland Thorns FC](https://www.nwslsoccer.com/team/portland-thorns-fc/stats),
a National Women’s Soccer League (NWSL) team.

In 2021, the Thorns won the:

+ NWSL Challenge Cup
+ Women’s International Champions Cup
+ NWSL Shield

The data come from the
[nwslR package](https://github.com/adror1/nwslR),
which provides tools to scrape and organize publicly available NWSL data.

The following functions pull different types of data:

+ `load_matches()` – match-level data
+ `load_players()` – player-level data
+ `load_teams()` – team information
+ `load_metrics()` – metric definitions

Running the code below will load four datasets into your environment.

```{r}
matches <- load_matches()
players <- load_players()
teams   <- load_teams()
metrics <- load_metrics()
```

---

# Exercises

All plots should follow best visualization practices discussed in lecture, including clear titles, labeled axes, and thoughtful aesthetic choices.

All code should follow the
[tidyverse style guidelines](https://style.tidyverse.org/pipes.html).

```{r, echo = FALSE}
portland_matches <- matches |> 
  filter(home_team_id == 8 | away_team_id == 8)
```

## Data Wrangling

Before beginning the analysis, you will need to wrangle the data to focus only on Portland Thorns matches.

::: callout-note
### Exercise 1
+ Inspect the `teams` data to determine the `external_team_id` for the Portland Thorns.
+ Filter `matches` to keep only games in which Portland played (as the home or away team).
+ Save the result as `portland_matches`.

Confirm that the dataset has `r nrow(portland_matches)` observations before proceeding.
:::

To add team names and abbreviations, we need to join the `teams` data twice:

+ once for the home team
+ once for the away team

We want the following new variables:

- `home_team_name`
- `home_team_abbr`
- `away_team_name`
- `away_team_abbr`

The code below creates `home_team_info` and joins it to `portland_matches`.

```{r, eval = FALSE}
home_team_info <- teams |> 
  select(home_team_id = external_team_id,
         home_team_name = team_name,
         home_team_abbr = team_abbreviation)

portland_matches <- portland_matches |> 
  left_join(home_team_info)
```


```{r, echo = FALSE}
away_team_info <- teams |> 
  select(away_team_id = external_team_id,
         away_team_name = team_name,
         away_team_abbr = team_abbreviation)

portland_matches <- portland_matches |> 
  left_join(away_team_info)
```

::: callout-note
### Exercise 2
+ Include the code above in your .qmd.
+ Then create an analogous `away_team_info` dataset with an `away_` prefix and join it to `portland_matches`.

Confirm the dimensions are `r nrow(portland_matches)` × 19 before continuing.
:::

::: callout-note
### Exercise 3
Create three new variables:

- `POR_points`
- `OPP_points`
- `result` (Win, Loss, or Tie for Portland)

Use the scaffold below and fill in the blanks.

```{r, eval = FALSE}
___ <- ___ |> 
  ___(POR_points = if_else(home_team_abbr == "POR",
                           home_team_score, away_team_score),
      OPP_points = if_else(___ ___ ___, ___, ___),
      result = case_when(POR_points  >  OPP_points ~ "Win",
                         POR_points  <  OPP_points ~ "Loss",
                         POR_points == OPP_points ~ "Tie"))
```
:::

## Match Analysis

::: callout-note
### Exercise 4
+ Create a bar chart showing the distribution of `result`.
+ Include clear axis labels and a title.
+ Interpret the results.
:::

::: callout-note
### Exercise 5
+ Create a visualization examining Portland’s performance by season (wins, losses, and ties).

+ Describe 2–3 patterns you observe.
:::

::: callout-note
### Exercise 6
*Does Portland have a home-field advantage?*

To investigate this question, 

+ Create a variable indicating whether Portland was the home or away team
+ Make side-by-side boxplots of points scored
+ Create a ridge plot using `geom_density_ridges()`

Describe 2–3 observations.
:::

::: callout-note
### Exercise 7
+ Create a formatted summary table showing the proportion of games won by whether the game was home or away.
+ Do the Thorns perform better at home or away?
:::

## Player Stats

The following code loads match-level player statistics.
This step may take several minutes. 

```{r, eval = FALSE}
#| eval: false
safe_load_pms <- purrr::possibly(load_player_match_stats,
                                 otherwise = data.frame())

player_stats <- purrr::map_df(portland_matches$match_id,
                              safe_load_pms,
                              .progress = TRUE)
```

In order to avoid loading the data everytime you render, add `#| eval: false` to the beginning of the above code chunk. After it has loaded, you can save the `player_stats` data by running `saveRDS(player_stats, "data/player_stats.rds")`. Then, include the following code in a new code chunk:

```{r}
# Load saved version
# avoids lengthy re-loading from package on each Render
player_stats <- readRDS("./data/player_stats.rds")
```

The research questions:

+ Which Portland players took the highest **percentage of shots outside the box**?
+ Which Portland players **converted the highest percentage of shots**?

::: callout-note
### Exercise 8
This is one big open-ended exercise to answer the two research questions above. It’s up to you to determine the appropriate wrangling necessary to answer the questions. 

A few tips for getting started:

+ Examine `player_stats` using `View()` to determine what each row represents. Are there multiple rows per player? If so, what does each row represent?
+ You’ll only need the`player_id`, `team_id`, `shots_total`, `shots_outside_box`, and `goals` variables to answer these questions
+ Filter the data to only keep Portland players
+ Restrict your analysis to only players that took at least 15 shots across all matches
+ You will eventually need to join in info from the `players` data so that your end result will display the players names and not just their numerical id

**Your final result for each question should be either a nicely formatted table, or a nicely formatted visualization, like the ones below. Use these end results to help you decide what wrangling will help you build up to this information.**
:::

```{r, echo = FALSE, message = FALSE}
summary_data <- player_stats %>% 
  filter(team_id == 8) %>% 
  select(player_id, team_id, shots_total, shots_outside_box, goals) %>% 
  group_by(player_id) %>% 
  summarize(shots_total = sum(shots_total),
            shots_outside_box = sum(shots_outside_box, na.rm = TRUE),
            goals = sum(goals, na.rm = TRUE)) %>%
  filter(shots_total >= 15) %>% 
  mutate(prop_outside_box = shots_outside_box/shots_total,
         prop_goals = goals/shots_total) 

summary_data %>% 
  arrange(desc(prop_outside_box)) %>% 
  left_join(players) %>% 
  slice(1:10) %>% 
  ggplot(aes(x = prop_outside_box, 
             y = reorder(player_match_name, prop_outside_box))) +
  geom_col(fill = "purple") +
  theme_minimal() +
  scale_x_continuous(labels = scales::percent) +
  labs(x = "Percentage of shots taken outside the box",
       y = "",
       title = "Portland players with Highest % of shots taken outside the box",
       subtitle = "Minimum 15 shots")

summary_data %>% 
  arrange(desc(prop_goals)) %>% 
  left_join(players) %>% 
  slice(1:10) %>% 
  select(player_match_name, shots_total, goals, prop_goals) %>% 
  kable(format = "html", table.attr = "style='width:50%;'") %>% 
  kableExtra::kable_styling()
```


---

## Bonus (2 pts)

Propose an additional question you could investigate using these data.
Provide a visualization or table and briefly interpret the result.

---

➡️ ✅ *Render one last time and open `r-hw-04`.html` in a web browser. Check that your code, output, visualizations, and written responses are organized and readable.*

# Submit

::: {.callout-important}

To submit, you need to upload your R_HW_04.html file to the R HW 04 assignment on Brightspace. The html file should be located in the same folder on your computer where you placed your .qmd (STAT_7500). 

:::

# Grading (10 points)

Homework assignments are opportunities to practice the skills and habits you will use on larger course projects. They are graded holistically rather than point-by-point.

Your homework will be evaluated for:

- **Completion and substantive correctness (8 points):** You have attempted all exercises, your code and results are generally correct, and your written responses appropriately interpret and engage with the data.
- **Reproducible workflow and presentation (2 points):** Your document renders successfully and your code, output, visualizations, and responses are organized and readable.

Minor mistakes will not be penalized individually. You may also receive feedback about misconceptions or habits to improve even when they do not result in a point deduction. Use this feedback in future homework assignments and projects.

### Homework rubric — 10 points total

#### Completion and substantive correctness — 8 points

| Score | Description |
|---|---|
| **8** | Complete and substantively correct. Code and analyses accomplish what was asked, and written responses appropriately interpret the results. Minor imperfections are okay. |
| **7** | Mostly complete and correct, with one notable error, omission, or misunderstanding. |
| **6** | Several errors or omissions, but the work demonstrates reasonable understanding of the main ideas. |
| **4–5** | Significant errors, omissions, or misunderstandings that should be addressed before moving on. |
| **0–3** | Substantially incomplete, incorrect, or shows limited engagement with the homework. |

#### Reproducible workflow and presentation — 2 points

| Score | Description |
|---|---|
| **2** | Document renders successfully; code, output, visualizations, and written responses are reasonably organized and readable. |
| **1** | Some problems with reproducibility, organization, readability, or presentation. |
| **0** | Major problems with reproducibility, organization, readability, or presentation. |

**Total: 10 points**

