---
title: "R HW 03 - Data Visualization"
format:
  html:
    css: styles.css
    grid:
      margin-width: 350px
fig-numbering: false
toc: true
toc-location: right
categories:
  - Data visualization
  - ggplot2
  - ANES
editor_options:
  chunk_output_type: console
callout-icon: false
---

# Goals

The main goal of this assignment is to practice selecting, constructing, and interpreting visualizations that are appropriate for the variables and research questions being examined.

You will...

+ Investigate the source and collection of a real-world dataset
+ Inspect the structure and dimensions of a dataset
+ Visualize categorical and numerical variables
+ Compare counts and proportions in bar charts
+ Compare distributions using boxplots and ridge plots
+ Evaluate whether a visualization is appropriate for a particular question
+ Apply principles of effective data visualization
+ Interpret visualizations in context and identify limitations

The individual R commands matter, but the larger goal is to develop a reliable process: begin with a meaningful question, identify the relevant variables and their types, select an appropriate visualization, and interpret what the visualization does—and does not—show.

# Getting Started

To get started on this assignment, do the following:

- Open our class [GitHub repo](https://github.com/kgfitzgerald/STAT_7500_templates)
- Navigate to `R_HW_03.qmd`
- Click the download icon (hover text should say "Download raw file")
- Place the downloaded `.qmd` document in your STAT_7500 folder
- Download `anes2020_subset.csv` and place it inside the `data` folder in your STAT_7500 project
- From inside your STAT_7500 project, open and render the Quarto document

Your project should be organized approximately like this:

```         
STAT_7500/
├── R_HW_03.qmd
└── data/
    └── anes2020_subset.csv
```

The document you are currently reading contains all the instructions for R HW 03. You will complete the assignment by adding code and text responses to the `R_HW_03.qmd` document.

::: column-margin
Keeping the data and analysis files together in an R Project allows you to use a relative file path such as `"data/anes_2020_subset.csv"`. Do not use a file path that begins with a user name or refers to a location specific to your computer.
:::

## Packages

In this assignment, we will use **tidyverse** to work with and visualize the data, **viridis** for accessible color palettes, and **ggridges** to create ridge plots.

You can run `install.packages("nameofpackage")` in your Console if you do not already have a package installed. Then, load the packages in your Quarto document:

```{r}
#| label: load-packages
#| message: false
#| warning: false

library(tidyverse)
library(viridis)
library(ggridges)
```

# Data: 2020 American National Election Study

The data for this homework assignment come from the [2020 American National Election Study](https://electionstudies.org/data-center/2020-time-series-study/).

Import the data using the code below:

```{r}
#| label: import-anes
#| warning: false
#| message: false

anes <- read_csv("data/anes2020_subset.csv")
```

The dataset contains the following variables:

- `CASEID`: a unique identification number for each respondent
- `hunt_fish`: whether the respondent went hunting or fishing in the past year
- `scientists`: a feeling-thermometer rating indicating how warmly the respondent feels toward scientists, ranging from 0 (coolest) to 100 (warmest)
- `education`: an ordinal categorical variable representing the respondent's highest level of education, ranging from less than high school to a professional degree
- `ideology`: a seven-point self-rating scale ranging from most liberal to most conservative
- `urbanrural`: the type of community in which the respondent lives, with the possible values rural, small town, suburb, and city

# Exercises

All plots should follow the visualization practices discussed in class. Unless there is a reason to do otherwise, include an informative title, descriptive axis labels, and a clear legend when needed. Make purposeful aesthetic choices that support the question being investigated.

Your written responses should interpret the visualization in context rather than merely restating what appears on the axes.

## Understanding the data

::: callout-note
### Exercise 1

Use the [ANES website](https://electionstudies.org/data-center/2020-time-series-study/) to learn more about the study.

Briefly describe:

+ who collected the data;
+ when the data were collected;
+ who was surveyed; and
+ anything else about the data-collection process that seems relevant or important for interpreting the results.
:::

::: callout-note
### Exercise 2

Inspect the `anes` dataset using appropriate R functions.

+ How many rows are in the dataset?
+ How many columns are in the dataset?
+ What does one row represent?

Include code and output to support your response.
:::

```{r}
#| label: inspect-anes

```

`r emo::ji("pause_button")` `r emo::ji("right_arrow")` ✅ *This is a good place to pause and Render to make sure there are no errors in your .qmd.*

## Visualizing one categorical variable

::: callout-note
### Exercise 3

**How do respondents describe their political ideology?**

Create a bar chart displaying respondents' `ideology`, with the number of respondents on the y-axis.

Then, respond to the following:

+ What is the most common ideology?
+ Do respondents tend to be moderate or more ideologically extreme?
+ Are there any missing or nonresponse categories that should be considered when interpreting the visualization?
:::

```{r}
#| label: ideology-bar-chart
#| fig-fullwidth: true

```

## Comparing two categorical variables

::: callout-note
### Exercise 4

**Does the type of community in which respondents live differ according to their political ideology?**

Create a filled bar chart with:

+ one bar for each level of `ideology`;
+ `urbanrural` mapped to the fill aesthetic; and
+ each bar scaled to a total height of 1 so that the segments represent proportions.

You are encouraged, but not required, to use a **viridis** color palette. Include clear labels for the axes, legend, and plot.

Then, respond to the following:

+ Where do respondents with different political ideologies tend to live?
+ Which comparisons are made easier by displaying proportions rather than counts?
+ Does the proportion of nonresponses vary substantially by ideology?
:::

```{r}
#| label: ideology-urban-rural
#| fig-fullwidth: true

```

`r emo::ji("pause_button")` `r emo::ji("right_arrow")` ✅ *This is a good place to pause and Render to make sure there are no errors in your .qmd.*

## Visualizing one numerical variable

::: callout-note
### Exercise 5

**How warmly do respondents feel toward scientists?**

Create a histogram with:

+ the feeling-thermometer rating (`scientists`) on the x-axis; and
+ the number of respondents on the y-axis.

Experiment with the number or width of the bins and select a version that communicates the distribution clearly.

Describe the distribution, including its shape, peaks, spread, and any unusual features. Explain how your choice of bins affects what is visible.
:::

```{r}
#| label: scientists-histogram
#| fig-fullwidth: true

```

## Comparing distributions

::: callout-note
### Exercise 6

**Does the political ideology of respondents who went hunting or fishing in the past year differ from the ideology of those who did not?**

Create side-by-side boxplots comparing the two groups.

Begin with the following code, which removes observations with missing values for `hunt_fish` and creates clearer labels for the two groups:

```r
anes |>
  drop_na(hunt_fish) |>
  mutate(
    hunted_fished = if_else(
      hunt_fish == 0,
      "Did Not Hunt or Fish",
      "Hunted or Fished"
    )
  ) |>
```

Complete the pipeline by adding the code needed to construct and label the boxplots.
:::

```{r}
#| label: hunting-boxplots
#| fig-fullwidth: true

```

::: callout-note
### Exercise 7

Explore the same question as in Exercise 6 using `geom_density_ridges()` to construct a ridge plot.

Use the same two groups and make sure the plot is clearly labeled.
:::

```{r}
#| label: hunting-ridge-plots
#| fig-fullwidth: true

```

::: column-margin
You can learn more about ridge plots in the [Introduction to ggridges](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html).
:::

::: callout-note
### Exercise 8

Compare the boxplots and ridge plots from Exercises 6 and 7.

+ What do the two visualizations suggest about differences in ideology between the groups?
+ What is easier to see in the boxplots?
+ What is easier to see in the ridge plots?
+ Which visualization would you use to communicate the comparison? Briefly justify your choice.
:::

`r emo::ji("pause_button")` `r emo::ji("right_arrow")` ✅ *This is a good place to pause and Render to make sure there are no errors in your .qmd.*

## Evaluating a visualization

::: callout-note
### Exercise 9

**Is there a relationship between a respondent's education level and how they view scientists?**

Create a scatterplot with:

+ `education` on the x-axis;
+ `scientists` on the y-axis; and
+ a line of best fit created with `geom_smooth(method = "lm")`.

Then, evaluate the resulting visualization:

+ Is a scatterplot an especially useful way to display these variables? Why or why not?
+ How does the way `education` is measured affect the appearance and interpretation of the plot?
+ Does the fitted linear relationship make sense for these data?
+ What alternative visualization might communicate the relationship more effectively?

You do not need to create the alternative visualization unless you would like to explore it.
:::

```{r}
#| label: education-scientists
#| fig-fullwidth: true

```

## Open exploration

::: callout-note
### Exercise 10

Develop one additional question that can be investigated using the variables in the `anes` dataset.

Then:

+ Clearly state your question.
+ Identify the variables needed to investigate it.
+ Create an appropriate visualization.
+ Explain why the selected plot fits the variables and question.
+ Describe what the visualization reveals.
:::

```{r}
#| label: open-exploration
#| fig-fullwidth: true

```

## Synthesis

::: callout-note
### Exercise 11

Write a brief paragraph that synthesizes what you learned from your analyses.

Your paragraph should address:

+ the most important or interesting findings;
+ limitations of the data or analyses;
+ conclusions that cannot be supported by these visualizations alone; and
+ additional data that would help you investigate related questions.
:::

`r emo::ji("right_arrow")` ✅ ⬆️ *Yay, you're done! Render one last time and verify that everything is displayed correctly in your `R_HW_03.html` file. Open the file in a web browser and inspect the complete document before submitting.*

# Submit

::: {.callout-important}

To submit, you need to upload your R_HW_03.html file to the R HW 03 assignment on Brightspace. The html file should be located in the same folder on your computer where you placed your .qmd (STAT_7500). 

:::

# Grading (10 points)

Homework assignments are opportunities to practice the skills and habits you will use on larger course projects. They are graded holistically rather than point-by-point.

Your homework will be evaluated for:

- **Completion and substantive correctness (8 points):** You have attempted all exercises, your code and results are generally correct, and your written responses appropriately interpret and engage with the data.
- **Reproducible workflow and presentation (2 points):** Your document renders successfully and your code, output, visualizations, and responses are organized and readable.

Minor mistakes will not be penalized individually. You may also receive feedback about misconceptions or habits to improve even when they do not result in a point deduction. Use this feedback in future homework assignments and projects.

### Homework rubric — 10 points total

#### Completion and substantive correctness — 8 points

| Score | Description |
|---|---|
| **8** | Complete and substantively correct. Code and analyses accomplish what was asked, and written responses appropriately interpret the results. Minor imperfections are okay. |
| **7** | Mostly complete and correct, with one notable error, omission, or misunderstanding. |
| **6** | Several errors or omissions, but the work demonstrates reasonable understanding of the main ideas. |
| **4–5** | Significant errors, omissions, or misunderstandings that should be addressed before moving on. |
| **0–3** | Substantially incomplete, incorrect, or shows limited engagement with the homework. |

#### Reproducible workflow and presentation — 2 points

| Score | Description |
|---|---|
| **2** | Document renders successfully; code, output, visualizations, and written responses are reasonably organized and readable. |
| **1** | Some problems with reproducibility, organization, readability, or presentation. |
| **0** | Major problems with reproducibility, organization, readability, or presentation. |

**Total: 10 points**

