---
title: "R Mini Project 02"
subtitle: "Data Wrangling & Scraping"
date: "due Wednesday December 17, 11:59pm"
output: 
  tufte::tufte_html:
    css: "lab.css"
    tufte_variant: "envisioned"
    highlight: pygments
    toc: true
    toc_depth: 1
link-citations: yes
editor_options: 
  markdown: 
    wrap: sentence
---

# Project ground rules

- You should use the `tidyverse` package in R to complete all tasks. You are welcome to use other packages as well, but all tasks must be done withiin R. 
-   This is an INDIVIDUAL PROJECT. You may NOT work together or consult other people on this project. You may ask the instructor for clarification on what a particular exercise is asking, but the instructor will not provide guidance on how to complete the exercise or troubleshoot your code. 
-   You are ENCOURAGED to use your notes, lecture slides/videos, previous assignments, and the internet.
-   You may use Google or ChatGPT (or other LLM) for coding questions, but please do not ask them to answer the question itself. 
- If you use AI, include an Appendix section at the end, indicating which problems you used it on and the corresponding prompts you used. 
- You should submit your pdf on Blackboard as a **single well-documented rendered file**. 
- There is no specific length to the report, but you should exclude unnecessary code and output.  (In particular, please do not print the entire dataset or a large subset thereof). Use headers and subheaders to clearly delineate which exercise the code/output/commentary corresponds to 

# Goals

This Project will assess the following skills:

-   Data wrangling
-   Data visualization
-   Web scraping
-   Clarity of code and written communication
-   Ability to use data to investigate questions

# Getting started

+ You are not provided with a template .qmd for this assignment but should rather create your own. I recommend copy/pasting the YAML from a previous assignment to start with similar formatting. 
+ Check out the documentation for the [Fatal Force Database](https://github.com/washingtonpost/data-police-shootings/tree/master) hosted on GitHub by the Washington Post. The Washington Post compiles a database of every fatal shooting in the United States by a police officer in the line of duty since January 1, 2015.  The Post is tracking more than a dozen details about each killing — including the race of the deceased, the circumstances of the shooting, and whether the person was armed — by culling local news reports and monitoring independent databases, such as Killed by Police and Fatal Encounters. In some cases, The Post conducted additional reporting.  The Post is documenting only shootings in which a police officer, while on duty, shot and killed a civilian — circumstances that most closely parallel the 2014 killing of Michael Brown in Ferguson, Mo. The Post is not tracking deaths of people in custody, fatal shootings by off-duty officers, or deaths in which police gunfire did not kill the individual.  The FBI and the Centers for Disease Control and Prevention log fatal shootings by police, but officials acknowledge that their data is incomplete. 
+ Download version 2 of the fatal-police-shootings-data.csv data, which can be found [here](https://github.com/washingtonpost/data-police-shootings/blob/master/v2/fatal-police-shootings-data.csv), and read the data into R using `read_csv()`


```{r setup, message = FALSE, echo = FALSE}
library(tidyverse)
library(here)
```

```{r, echo = FALSE, warning=FALSE, message = FALSE}
shootings <- read_csv(here("assets/homework/R_HW_instructions/data/fatal-police-shootings-data.csv"))
```

# Analyses to complete

For all exercises, you are welcome to go beyond the specific request of the exercise. If there are additional analyses or visualizations that help you (and the reader) gain further insight into the story the data is telling, feel free to include them. Remember to add appropriate labels for all visualizations and provide written commentary on what insight can be gleaned from each output. 

1. **Univariate Summaries** Produce an appropriate graphical display for the following variables, each one by itself and describe the distribution of the data.
    + `date`
    + `age`
    + `gender`
    + `race`
    + `city`
    + `state`
    + `body_camera`

> *Note 1 - You are welcome (and encouraged) to create a visualization with a subset of the data or with the data modified in some way to get a better picture of the distribution.*

> *Note 2 – Make sure to explore and discuss the missing values in each variable as well.*

*If you already completed any of Exercises 2-4 in your work for Exercise 1, please do not leave it blank, but rather add a comment in the part below that says something like “see Exercise 1 above”. In practice, it’s a good idea to do a univariate summary for all variables, but I am confident that if you can do it for these variables, you can do it just as well for the others.*

2. Using the `date` variable, create visualizations for the i) day of the week and ii) year of the shootings.  These variables can be derived from the date variable (*Hint: use lubridate’s `year` and `wday` functions along with the `label=T` argument for day of the week*).
3. Recreate the visualization for the `body_camera` variable, but change the labels to “Yes” instead of TRUE and “No” instead of FALSE.  (You should do this moving forward for this variable, as well as for `was_mental_illness_related`, if you use them again)
4. Recreate the visualization for the `race` variable, but change the categories into White (W), Black (B), Hispanic (H), and Other (all others, including O).  Make sure to leave missing values as missing values. Use `geom_text` to add count labels at the end of each bar (*you may want to ask Google or ChatGPT how to do this if you haven't done it before*) 

> *Note, you should use this simplified version of the race variable for all future analyses involving race*

5. Produce visualizations using each of the following *pairs* of variables (one graph with two aesthetics)
    + `body_camera` and `race` - The goal here is to compare the percent of times a body camera was used and compare across races.
    + `age` and `race` – explore the age distribution across different racial groups
    + `race` and `year` – explore the distribution of race over time (by year)
6. Produce a visualization that uses three variables, one of which should be `was_mental_illness_related`.  Explain why you chose the variables and type of graph that you did.  Describe what you see in the visualization.
7. Using the shooting data, create a dataset where the case/observational unit is a state (one row per state) with the following derived variables.  For each one, make sure to deal with missing values appropriately.  For each variable, graph the values in descending order (Hint: the reorder function is a good choice here).  Note, you do not need to include all states on all graphs, but the ones that you choose to show should be listed in descending order (for example, you could show only the top 10 in order or the top 5 in order along with the bottom 5 in order).
    + The number of shootings in that state.
    + The median age of people shot in that state.
    + The proportion of victims that were White.
    + The proportion of victims who were fleeing (based on `flee_status`, which includes all values except “not” but should exclude NAs).
8. Using the by-state data you derived above, create a scatterplot of the proportion of victims that were White vs. the proportion of victims who were fleeing. In this graph, you should find a single outlier (influential point).
    + Label this outlier with the name of the state.
    + Remove the outlier from the graph.  State whether you think this is an appropriate choice.
    
## Data Scraping

We are interested in scaling the data to include population and demographic information.  

Population per state can be found at [https://simple.wikipedia.org/wiki/List_of_U.S._states_by_population](https://simple.wikipedia.org/wiki/List_of_U.S._states_by_population).  

“Racial breakdown of population by state” can be found at [https://en.wikipedia.org/wiki/Demography_of_the_United_States](https://en.wikipedia.org/wiki/Demography_of_the_United_States) (almost half the way down the page).

9. Read the above tables into a data frame (or tibble) in R using the `rvest` package.

> Note - If you can’t scrape the data using `rvest`, then you can copy and paste the tables into Excel and read them in that way or use some other method of your choosing.  If you do not scrape them using `rvest`, you will not get credit for this part, but it will enable you to complete the sections below.

10. Calculate the number of shootings per million people living in the state.  Put these onto a graph in descending order.  Note, you do not need to include all states on the graph. How does this graph compare to the first one you created in Exercise 7 above?
11. Create a scatterplot of the proportion of victims who are White vs. proportion of the population that is White.  You should have one case (point) per state.  Comment on what you see.


# Submission

Make sure to inspect your rendered html for neatness, clarity, and completeness of all parts of each exercise. Once you are satisfied with your final report, save it as a pdf and submit it on Blackboard. 
