---
title: "R Mini Project 01"
subtitle: "Data Wrangling & Visualization"
date: "due Tuesday, November 25, 11:59pm"
output: 
  tufte::tufte_html:
    css: "lab.css"
    tufte_variant: "envisioned"
    highlight: pygments
    toc: true
    toc_depth: 1
link-citations: yes
editor_options: 
  markdown: 
    wrap: sentence
---

# Project ground rules

-   This is an INDIVIDUAL PROJECT. You may NOT work together or consult other people on this project.
-   You are ENCOURAGED to use your notes, lecture slides/videos, previous assignments, and the internet.
-   You may use Google or ChatGPT (or other LLM) for coding questions, but please do not ask them to answer the question itself. 
- If you use AI, include an Appendix section at the end, indicating which problems you used it on and the corresponding prompts you used. 

# Goals

This Project will assess the following skills:

-   Data wrangling
-   Data visualization
-   Clarity of code and written communication
-   Ability to use data to investigate questions

# Getting started

-   Download `R-MP-01.qmd`, `avengers.csv` and `World Happiness 2023.xlsx` from the course website
-   Place the .qmd file in your `STAT_7500` folder on your computer
-   Place the .csv and .xslx data files inside the `data` subfolder within `STAT_7500`

# Packages

We will use the following packages for this project:

```{r setup, message = FALSE}
library(tidyverse)
library(kableExtra)
library(readxl)
```

```{r, echo = FALSE, eval = FALSE}
happiness <- read_excel("assets/homework/R_HW_instructions/data/World Happiness 2023.xlsx")

happiness |> 
  group_by(year) |> 
  summarize(sum(is.na(logGDPpc)))
```


# Data

You will analyze two different datasets in this project:

-   `avengers` (Exercises 1 - 5): data on 173 Marvel characters
-   `happiness` (Exercises 6 - 9): data on World Happiness metrics for 165 countries from 2008 to 2023

# Avengers Data

This data was originally collected for [a FiveThirtyEight article](https://fivethirtyeight.com/features/avengers-death-comics-age-of-ultron/).
The version of the `avengers` data we will work with here can be found in the `avengers.csv` file on the course website.
The code below will load the data.

```{r, eval = FALSE}
avengers <- read_csv("data/avengers.csv")
```

This dataset includes information about characters across the entire Marvel Cinematic Universe (MCU), so some of the names will be familiar if you are a fan of the films or comics.
Don't worry if you aren't a Marvel fan; no background knowledge is needed to successfully complete this project!

We will focus on the following variables in this project:

| Header                | Definition                                                                                                                                                                        |
|---------------------|---------------------------------------------------|
| `name`                | The full name or alias of the character                                                                                                                                           |
| `appearances`         | The number of comic books that character appeared in as of April 30                                                                                                               |
| `current`             | Is the member currently active on an avengers affiliated team?                                                                                                                    |
| `gender`              | The recorded gender of the character                                                                                                                                              |
| `probationary_introl` | Sometimes the character was given probationary status as an Avenger, this is the date that happened. The value will be *NA* if the character was never given probationary status. |
| `full_reserve`        | The month and year the character was introduced as a full or reserve member of the Avengers                                                                                       |
| `year`                | The year the character was introduced as a full or reserve member of the Avengers                                                                                                 |
| `years_since_joining` | 2015 minus the year                                                                                                                                                               |
| `death1`              | Yes if the Avenger died, No if not.                                                                                                                                               |
| `return1`             | Yes if the Avenger returned from their first death, No if they did not, blank if not applicable                                                                                   |

See FiveThirtyEight's [GitHub repo](https://github.com/fivethirtyeight/data/tree/master/avengers) for the full codebook.

1.  You are interested in creating a data frame with the most "classic" Avengers.
    Create a new data frame that only includes Avengers that were 1) created in 1970 or earlier and 2) were NOT given probationary status.
    Call the new data frame `classic_avengers`.
    *Hint: `is.na()` will be useful. Check: your new data frame should have 27 observations.*

2.  *Who are the newest classic Avengers?* Using a single pipeline: Create a new variable called `years_served` that represents the number of years served **as of 2025.** *(Hint: you can use either the `year` variable or `years_since_joining` variables to do this.)* 
    
    - Arrange the data appropriately to display the newest classic Avengers first.
    - Display only necessary columns and the first 10 rows in a nicely formatted table.
    -   In addition to the table, report the names of the three newest classic Avengers and how long they have served in your narrative text.

3.  *Has the percentage of female Avengers changed over time?* To explore this question, compute the percentage of female Avengers in the `classic_avengers` dataset and compare it to the percentage of females among **all** Avengers.
    What do you conclude based on these results?

4.  Sort the full `avengers` dataset in descending order of `appearances` and display only the columns `name`, `appearances`, `death1`, and `return1` for the top five observations.

    -   What do you observe about these Avengers in terms of deaths and returns?

5.  *Do characters who die at least once tend to appear more or less often than characters who never die?* Create a visualization using the full `avengers` dataset to examine the distribution of appearances by whether or not character has died at least once.
    There are multiple correct way to visualize this, just make sure your visualization is clean and easy to read.

    -   What do you learn about Marvel movies from your results?

# World Happiness Data

The World Happiness Report is produced annually by the Gallup World Poll. According to their website, 

> "Since creating the World Poll in 2005, Gallup has conducted studies in more than 160 countries and territories that are home to more than 98% of the world’s adult population. The World Poll survey includes more than 100 global questions as well as region-specific items. Gallup asks residents from Australia to Zimbabwe the same questions, every time, in the same way. This makes it possible to trend data from year to year and make direct country comparisons." - [World Happiness Report](https://www.worldhappiness.report/data-sharing/)

The `World Happiness 2023.xlx` data file includes data from 2008 to 2023. The following code will read it in. 

```{r, eval = FALSE}
happiness <- read_excel("data/World Happiness 2023.xlsx")
```


| Variable           | Description or Question(s) Asked                                                                                                                                                                                                                                                                                                                                                          |
|---------------------|--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| country             | Country name (used only as an identifier).                                                                                                                                                                                                                                                                                                                                                |
| year                | 2008 – 2023.                                                                                                                                                                                                                                                                                                                                                                              |
| happiness           | Country average response to “Imagine a ladder, with steps numbered from 0 at the bottom to 10 at the top. The top of the ladder represents the best possible life for you and the bottom of the ladder represents the worst possible life for you. On which step of the ladder would you say you personally feel you stand at this time?”                                              |
| logGDPpc            | log₁₀ of Gross Domestic Product Per Capita (in 2011 US$).                                                                                                                                                                                                                                                                                                                                 |
| social_support      | Having someone to count on in times of trouble. Proportion of people in that country who responded “yes” to “If you were in trouble, do you have relatives or friends you can count on to help you whenever you need them, or not?”                                                                                                                |
| life_exp            | Life expectancy, in years, of a healthy child at birth.                                                                                                                                                                                                                                                                                                                                   |
| freedom_choices     | Proportion of people in that country who responded “satisfied” to “Are you satisfied or dissatisfied with your freedom to choose what you do with your life?”                                                                                                                                                                                                                            |
| freedom_choices_cat | A categorization of `freedom_choices` into “low”, “med”, or “high”.                                                                                                                                                                                                                                                                                                                       |
| generosity          | Residual of national average of response to “Have you donated money to a charity in the past month?” on GDP per capita.                                                                                                                                                                                                                                                                   |
| corruption          | Average proportion of people responding “yes” to “Is corruption widespread throughout the government or not?” and “Is corruption widespread within businesses or not?”.                                                                                                                                                                                                                   |
| affect_pos          | Average of three questions: “Did you experience happiness, laughter, and enjoyment during a lot of the day yesterday?”                                                                                                                                                                                                                                                                     |
| affect_neg          | Average of three questions: “Did you experience worry, sadness, and anger during a lot of the day yesterday?”                                                                                                                                                                                                                                                                              |



6. Filter the data to keep only the year 2023. For each of the following variables (by themselves), produce an appropriate visualization: `happiness`, `logGDP`, `life_exp`, `social_support`, `freedom_choices_cat`, `corruption`

7. Using the 2023 data, produce graphs of happiness as the dependent (y) variable and each other variable from Exercise 6 as the independent (x) variables (one at a time; 5 total visualizations). For scatterplots, include a smoothed curve (geom_smooth) as well. 

    - Which variables appear to be most correlated with happiness? Explain your answer.

8.  Which countries had the top 5 highest happiness ratings in 2023? The lowest 5? Use code to identify these 10 countries and produce a nicely formatted table of the country names, along with their happiness ratings, life expectancy, corruption, and logGDPpc. 

```{r, echo = FALSE, eval = FALSE}
top_5 <- hap_2023 |> 
  slice_max(happiness, n = 5) |> 
  pull(country)

bottom_5 <- hap_2023 |> 
  slice_min(happiness, n = 5) |> 
  pull(country)

hap_2023 |> 
  filter(country %in% top_5 | country %in% bottom_5) |> 
  select(country, happiness, life_exp, corruption) |> 
  arrange(happiness)
```


9.  Using the 2023 data, create a new variable called `social_support_cat` which has the value "high" if the value of that country’s `social_support` is above the median and 0 if it is not. Create an appropriate visualization of this new variable by itself. Then, create a scatterplot of `happiness` as the dependent (y) variable and `logGDPpc` as the independent (x) variables, with `social_support_cat` as the color aesthetic.

10. Choose at least three, but no more than about eight, countries from the data that are of interest to you.

    + Explain why you chose those countries. (This part will not be graded – I am just curious)
    + For these countries, use only the 2023 data to produce at least a graph with happiness as the dependent (y) variable and another variable of your choice as the independent (x) variable, using a color aesthetic for each country.
    + Using all years of data, produce a graph of happiness over time, with a different color for each country you chose.

# Further exploration

11.  Choose one of the two datasets - `avengers` or `happiness` - and comment on each of the following:

-   What you found through this analysis

-   What limitations might exist to the data/analyses

-   What additional data you would like to have to explore further questions

# BONUS

State an additional research question you could investigate using either the `avengers` or `happiness` data. State the question, provide a visualization that investigates it, and comment on what your visualization shows.

# Submission

Before submitting your assignment on Blackboard, Render one last time, then open up your .html in a full window and inspect the following:

+ Check your code for neatness - add spaces and line breaks where appropriate to improve readability

+ Check any visualizations for informative and clean labels and titles

+ Check any tables for clean formatting (e.g. with `kbl() |> kable_minmal()`)

+ Check that you've interpreted output as appropriate

+ Make sure extraneous warnings and messages are suppressed (e.g. set `#| warning: false`, `#| message: false`)

+ Check that Exercises are clearly labeled and your text responses are visually distinguished

+ Check for overall neatness & organization of the report (e.g. headers, subheaders, use of bullets, font changes)

To submit, export your html as a .pdf and upload the pdf document to the relevant assignment on Blackboard. Refer back to R HW 01 for detailed instructions on how to submit your assignment. 


# Grading (100 pts)

<br>

| Component          | Points |
|:-------------------|:-------|
| Ex 1               | 8      |
| Ex 2               | 12      |
| Ex 3               | 8      |
| Ex 4               | 6      |
| Ex 5               | 12      |
| Ex 6               | 12      |
| Ex 7               | 10      |
| Ex 8               | 8      |
| Ex 9               | 8      |
| Ex 10 | 8      |
| Ex 11 | 8      |
| BONUS              | 2      |
