---
title: "R HW 04: Data Wrangling II"
subtitle: "Women's Soccer"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
format: 
  html:
    self-contained: true
    toc: true
    toc_float: true
    number_section: false
    highlight: "tango"
    theme: "cosmo"
editor: visual
editor_options: 
  chunk_output_type: console
---


```{r}
#| label: load-packages
#| message: false
library(tidyverse)
library(viridis)
library(kableExtra)
library(ggridges)
library(scales)
#install.packages("devtools")
devtools::install_github("nwslR/nwslR", ref = "v0.0.0.9002")
library(nwslR)
```

```{r}
#| label: load-data
#| message: false
matches <- load_matches()
players <- load_players()
teams <- load_teams()
metrics <- load_metrics()
```

# Exercise 1

```{r}

```

# Exercise 2

```{r}

```

# Exercise 3

```{r}

```

# Exercise 4

```{r}

```

# Exercise 5

```{r}

```

# Exercise 6

```{r}

```

# Exercise 7

```{r}

```

# Exercise 8

```{r}

```

# BONUS

# Reflection prompts

*Respond to each of the following after you have completed all exercises in this assignment*

*(RP1): What were the main concepts covered in this assignment?*

> YOUR ANSWER HERE

*(RP2): What's one thing you understand better after completing these exercises?*

> YOUR ANSWER HERE

*(RP3): What exercise(s) gave you the most trouble? What was difficult about them/where did you get stuck?*

> YOUR ANSWER HERE