---
title: "R Mini Project I"
subtitle: "Avengers & World Happiness Data"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
format: 
  html:
    self-contained: true
    toc: true
    toc_float: true
    number_section: false
    highlight: "tango"
    theme: "cosmo"
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r setup}
#| label: load-packages
#| message: false
library(tidyverse)
library(kableExtra)
```

```{r}
#| label: load-data
#| message: false
avengers <- read_csv("data/avengers.csv")
happiness <- read_excel("data/World Happiness 2023.xlsx")
```

# Avengers analysis

## Exercise 1

```{r}

```

## Exercise 2

```{r}

```

> YOUR ANSWER HERE

## Exercise 3

```{r}

```

> YOUR ANSWER HERE

## Exercise 4

```{r}

```

> YOUR ANSWER HERE

## Exercise 5

```{r}

```

> YOUR ANSWER HERE

# World Happiness analysis

## Exercise 6

```{r}

```

## Exercise 7

```{r}

```

> YOUR ANSWER HERE

## Exercise 8

```{r}

```

## Exercise 9

```{r}

```

## Exercise 10

> YOUR ANSWER HERE

```{r}

```

## Exercise 11

> YOUR ANSWER HERE

# BONUS

# Appendix 

Use this section to indicate which problems you used AI for help (if any), and provide the corresponding prompts you gave it. If you did not use AI, simply state that here. 