---
title: "Misc Topics / Parting Thoughts"
subtitle: "STAT 7500"
author: "Katie Fitzgerald"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css", "slides.css"]
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r child = "setup.Rmd"}
```

```{r packages, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

# Let's recap / reflect

+ What's something you can do now that you couldn't do before this class?

--

+ Is there a way in which you *think* differently than you did at the beginning of this class? (e.g. "habits of mind", ways of approaching problems, data ethics considerations, other?)

---

# What package do I need?

---

# What function do I need?

[STAT 7500 Communal Cheat Sheet](https://docs.google.com/document/d/1K3Anyz4Qq79xUBjIkzSlQhjisbhu6cIPY-GMSp_zcq0/edit?usp=sharing)

+ Make a list of every function used 
+ Provide brief definition of what it's used for 
+ Look at slides, in-class .qmds, homework

---

class: middle

# Additional programming tools & resources

---

# Additional programming tasks

+ functions
+ for loops / iteration
+ simulations, random variables & distributions
+ statistical inference - base R, infer
+ modeling
+ visualization
+ Interactive Web apps / dashboards

---

# Writing Functions

Useful when you find yourself thinking:

+ "I'm repeating the same code over and over"
+ "Only one part keeps changing"
+ "If I make one improvement, I wish it applied everywhere"
--

```{r, eval = FALSE}
plot_histogram <- function(data, var) {
  ggplot(data, aes(x = .data[[var]])) +
    geom_histogram(bins = 30, color = "white") +
    labs(title = paste("Histogram of", var),
         x = var,
         y = "Count") +
    theme_minimal()
}
plot_histogram(penguins, "bill_length_mm")
```

---

# Writing Functions: Use cases

+ Create a histogram for every numeric variable
+ Produce a summary table for each STEM course
+ Write out a csv file of respondents for every professor/class
+ Other? 

---

# Writing Functions: Your Turn

How could this be turned into a function?

```{r, eval = FALSE}
(a - min(a, na.rm = TRUE)) / (max(a, na.rm = TRUE) - min(a, na.rm = TRUE))
(b - min(b, na.rm = TRUE)) / (max(b, na.rm = TRUE) - min(b, na.rm = TRUE))
(c - min(c, na.rm = TRUE)) / (max(c, na.rm = TRUE) - min(c, na.rm = TRUE))
(d - min(d, na.rm = TRUE)) / (max(d, na.rm = TRUE) - min(d, na.rm = TRUE))  
```


---

# Iteration / for loops / `purrr`

.pull-left[
"Do this iteratively, for every element"

+ Already learned some powerful tools that perform the same action on multiple things:

    + `facet_wrap`
    + `group_by` plus `summarize`

+ To do something to multiple columns, use `across()`
]

--

.pull-right[
```{r, eval = FALSE}
df |> summarize(
  n = n(),
  a = median(a),
  b = median(b),
  c = median(c),
  d = median(d),
)

df |> summarize(
  n = n(),
  across(a:d, median),
)
```
]

---

# For loops 

```{r}
squares <- numeric(5)  # create an empty numeric vector of length 5
for (i in 1:5) {
  squares[i] <- i^2
}
squares
```

--

BUT, R is vectorized, so it's quicker to just do 

```{r}
x <- 1:5
squares <- x^2
squares
```


---

# Iteration with `apply` family

```{r}
#lapply over a list
l <- list(a = 1:5, b = 6:10)
lapply(l, sum)

# sapply returns a vector/matrix
sapply(l, sum)
```

---

# Iteration with `purrr` package (`tidyverse`)

```{r}
nums <- list(1:5, 6:10)
map_dbl(nums, sum)
```

Use case: reading in a bunch of data files

```{r, eval = FALSE}
grade_files <- paste("./raw data/", 
                     dir(path = "./raw data/", 
                        pattern = "grade",
                        ignore.case = TRUE), sep = "")

grades <- grade_files %>% 
  map_dfr(read_excel, sheet = 2) %>%
  clean_names() %>%
  filter(term == 2237)
```

---

### Simulations / Random variables & Distributions

Why simulate?

+ helps us understand randomness / sampling variability
+ allows us to tackle questions we can't solve analytically
+ allow us to test methods / estimators (is this unbiased? which one has smaller variance?)
+ allows us to model / forecast potential outcomes, possible futures (e.g. in finance, public health, environmental risk)

---

### Simulations / Random variables & Distributions

+ `rnorm()`, `rpois()`, etc - generate random variables from a distribution
+ `sample()` - take a random sample of observations, with or without replacement
+ `sample_n()` to take random sample of rows in dplyr pipeline 
+ `replicate()` - concise way to repeat simulations

```{r}
replicate(1000, mean(rnorm(50)))
```

---

```{r}
sims <- data.frame(xbar = replicate(1000, mean(rnorm(50))))
ggplot(sims, aes(x = xbar)) +
  geom_histogram(color = "white")
```

---

# Statistical inference in R

+ `t.test()`, `prop.test()`, `chisq.test()`, `anova()`
+ `infer` package - tidyverse friendly 

---

# Modeling in R

+ `lm` for basic linear models
+ `glm` for generalized linear models (e.g. logistic)
+ `summary()` - for coefficients / model output
+ `predict()` - predictions for any fitted model
+ `broom` package for tidying model output
    + `tidy()`, `glance()`, `augment()`
+ `lme4` package for mixed-effects models
+ `survival` package for survival models
+ `tidymodels` package - modeling & machine learning workflows

---

# Upping your visualization game

+ [Themes to spice up visualizations with `ggplot2`](https://towardsdatascience.com/themes-to-spice-up-visualizations-with-ggplot2-3e275038dafa/)
--

+ Combine plots: [cowplot](https://cran.r-project.org/web/packages/cowplot/vignettes/introduction.html) & [patchwork](https://patchwork.data-imaginist.com)
--

+ [Fundamentals of Data Visualization book](https://clauswilke.com/dataviz/index.html)
--

+ [The R Graph Gallery](https://r-graph-gallery.com)
--

+ [Color blindness simulator](https://www.color-blindness.com/coblis-color-blindness-simulator/#google_vignette)
--

+ [The Science of Visual Data Communication: What Works](https://faculty.sites.iastate.edu/tesfatsi/archive/tesfatsi/ScienceOfVisualDataCommunication.FranconeriEtAl2021.pdf)
--

+ Take my data viz course next year :) 

---

# Interactive Web Apps / Dashboards

[Shiny](https://shiny.posit.co)

+ My use cases:
  + [MLB Player's WAR stats](https://kgfitzgerald.github.io/baylor_apu_score/baseball/mlb_prime_age/)
  + [Z-scores](https://kgfitz.shinyapps.io/zscores/)
  + [Calculate your final grade](https://kgfitz.shinyapps.io/calculate_final_grade/)
  + Dashboard of enrollment trends
  
+ [Shiny Gallery](https://shiny.posit.co/r/gallery/)


---

### Quarto / RStudio Infrastructure: what else can it do?

+ Slides
   + [Quarto website](https://quarto.org/docs/presentations/)
   + [Github repo](https://github.com/kgfitzgerald/stat-7500) for our [course website](https://kgfitzgerald.github.io/stat-7500-f25-archive/)
   + [Xaringan presentations](https://bookdown.org/yihui/rmarkdown/xaringan.html) (R Markdown)
--

+ Websites
    + Our course website - [Just the Docs template](https://github.com/just-the-docs/just-the-docs)
    + [Personal website](https://katie-fitzgerald.netlify.app)
    + [MATH 250: Data Analysis](https://apumath250.netlify.app)
--

+ Books
    + [Intro to Stats and Data Science](https://nustat.github.io/intro-stat-ds/)
+ See [Quarto Gallery](https://quarto.org/docs/gallery/)
---

# Github

+ [GitHub](https://github.com) - storing, sharing, collaborating on code

    + [Happy Git and GitHub for the useR](https://happygitwithr.com)
    + Clone people's repositories and see how they code / build things!
    + Build a public presence on GitHub 
    
---

# Communities to join

+ [American Statistical Association](https://www.amstat.org/membership/become-a-member/membership-options) (~$30 student membership)
+ [R Ladies](https://www.rladiesphilly.org)

---

# Statistics for social good resources

+ [Data4SDGs](https://www.data4sdgs.org)
+ [HRDAG](https://hrdag.org)
+ [Statistics Without Borders](https://www.statisticswithoutborders.org)
+ [DataKind](https://www.datakind.org)
+ [Health of Federal Statistical Agencies / Nation's Data is at Risk](https://www.amstat.org/policy-and-advocacy/assessing-the-health-of-the-principal-federal-statistical-agencies)

---

# Suggested reading

+ [Communicating with data: The Art of Writing for Data Science](https://communicating-with-data.github.io)
+ [Bernoulli's Fallacy: Statistical Illogic and the Crisis of Modern Science](https://cup.columbia.edu/book/bernoullis-fallacy/9780231199957/)
+ [How Data Happened: A History from the Age of Reason to the Age of Algorithms](https://www.amazon.com/How-Data-Happened-History-Algorithms/dp/1324006730)
+ [Race after technology: Abolitionist Tools for the New Jim Code](https://www.amazon.com/Race-After-Technology-Abolitionist-Tools/dp/1509526404/ref=sr_1_1?crid=27KPV1IGM64XT&dib=eyJ2IjoiMSJ9.X9ombf3wyAE9d7vCaXniRAzEZZ8Q7-hjrC0Z7LCIExhsbGb-m8JqB8Kq7hiUM7INbPV3y3LT9RJezP7C4jHeSFDxW-CZHVdeA0JlVJwu446VNHsJOjYM0-Z8MYFB6QVD1KFg_6JYTCwUoOb9S85kGyNFBEa5wKubwiJ0VLvhDJ903H-OEIRfrxzepVYBCqyD1oVV0vkuZIR-msjRY2hHEA.RP9Hh9EpRRR7unLea4UiVQqPOVHhya7xnuxCcyWa5x8&dib_tag=se&keywords=race+after+technology&qid=1765394654&sprefix=race+after+tech%2Caps%2C117&sr=8-1)
+ [Automating Inequality: How High-Tech Tools Profile, Police, and Punish the Poor](https://www.amazon.com/Automating-Inequality-High-Tech-Profile-Police/dp/1250215781/ref=sr_1_1?crid=3TFDSG3BLH17J&dib=eyJ2IjoiMSJ9.LPoZC9frqeR5DSJFh2QBLAasy_6p3a_TEcMvWpeHmrsnBB5gSVaf2vj8OtihIHdIk4dbLJyH5doSdTSNZaPe7xPeH_7gT1CWsD71qSNweFUm9ghdtP-770ZhyuCEsdQw0renJPbLeglglgrwh8s20w.taOIpS93Nvy6PRFiLbx_lyAfZcxq3PxTbHMMc0UoeWA&dib_tag=se&keywords=automating+inequality&qid=1765394681&sprefix=automating+inequality%2Caps%2C114&sr=8-1)
+ [Invisible Women: Data Bias in a World Designed for Men](https://www.amazon.com/Invisible-Women-Data-World-Designed/dp/1419735217/ref=sr_1_1?crid=RSXPKTJBQF6K&dib=eyJ2IjoiMSJ9.boV-evutYwZevi0ZYJWYZyg3JsnXcZU-0rwqBHh9TTVihBqrQKXINYrqOtABPFRGwtEGVrH5STZvdbSGeKxLTdm3D2nedokivo7tN9swxCE7vQUJ_QQ0i4Fo8krUr8hsot-v8jjHm3C3rZhXr_yLPE1EhZVzuoUqbmj7kq6d95YOrX3jY6mfSHuXbVKAnPusy-ZEcc2rIRBwMdD6jeQQnQ0AO-JrkkB6b8BBe5Z5sfI.nwPOn69smpNYZlUYnsALebUoq_DlAiSdxMhdaod8rtE&dib_tag=se&keywords=invisible+women&qid=1765394713&sprefix=invisible+women%2Caps%2C135&sr=8-1)
+ [Mathematics for Human Flourishing](https://www.amazon.com/Mathematics-Human-Flourishing-Francis/dp/0300258518/ref=sr_1_1?crid=1QVCE98335WL&dib=eyJ2IjoiMSJ9.ZKJG4AtPX51kQ2BqN9c3Fl9CqYm2n7PfBoCdJ3E1VPI.cYdtlRDZUvpTKDjJug3w2C8n3Rupj0qE9sTEYW7grso&dib_tag=se&keywords=mathematics+for+human+flourishing+by+francis+su&qid=1765394733&sprefix=mathematics+for+hum%2Caps%2C131&sr=8-1)
+ [The Book of Why: The New Science of Cause and Effect](https://www.amazon.com/Book-Why-Science-Cause-Effect/dp/1541698967/ref=sr_1_1?crid=38721B91XRHZA&dib=eyJ2IjoiMSJ9.77Kcrbw5e-DGGNNcwPR_yzl30Z2ehwG3XR2OGxrxDvD8z8k5PaL9bysdJ2ySXagI9urFzTigt8FAUhF9K0rIRS6LgI6NNtBiB1xEqX_q4Ujxre-M7zTtuZCp3yBM9n_d5_4bh64SVKvvE1E6AodA87CjckaSIu_0v5Aijp3dWgdQ33phbJ9lSACeJAGJXK1FiZTfl44PTCEz4aKDWXKTJ7ucMp5zru8Arm827ClbZCQ.Iq8uipo4x3PzTYB4eWvyyTKehDRAR_yoSGunny6Mhz4&dib_tag=se&keywords=the+book+of+why&qid=1765394750&sprefix=the+book+of+why%2Caps%2C117&sr=8-1)
+ [The Lady Tasting Tea: How Statistics Revolutionized Science in the Twentieth Century](https://www.amazon.com/Lady-Tasting-Tea-Statistics-Revolutionized/dp/0805071342/ref=sr_1_1?crid=DXPTU7TMPAAP&dib=eyJ2IjoiMSJ9.avHxPviBBZwwDThTorkSU237idVigC6gg5K4V3XmCvS1m2tuWOJccP1XuzJfl9e3Phc9xfI3677ztfAdn9qNPfnpQqyuVpJu5dzWuTiQSHW7BoYu2EEiKLharYsSI2DsXTwqSJtzDo3a8f_JhjyuYnnmxuHLGpQ-2MXiqWOY0NZz2qug6dqh37qk9-25gxUOHvOWcI5765AU3Od6Hi2bGDL0uChsbwVrOaTq-ZKxiMo.tnICf2BFsVwtsr5VfyZAkryQOeGCHnSjB7Rr5wbPP34&dib_tag=se&keywords=the+lady+tasting+tea&qid=1765394770&sprefix=the+lady+tast%2Caps%2C122&sr=8-1)
+ [Thicker than Blood: How Racial Statistics Lie](https://www.amazon.com/Thicker-Than-Blood-Racial-Statistics/dp/0816639094/ref=sr_1_1?crid=QJZDL23TK86C&dib=eyJ2IjoiMSJ9.DHq92bbRD63ch9FSl6s-upc2huynGsD-FnMMQ8oGkWXMJpml25UsNSuXEL2lvzQs7xWH2VUww24p7e1MdBi6vhDzul6E00uAT70mo3ZprItslR1hTgUvxO7XM27oRh69qZIJBsafsnFWcblrnZQPQD5uS_RnBA1gLcWA6AdC8NVm15LPwnw64La49nKPzyLHHXiGqdJnxfptd0VAupO0NLYv37IuU6anRYKZV2MzV1w.SssIAE8yIKkuVaMayzCk-h8KKsc34iXVpdx5499WEic&dib_tag=se&keywords=thicker+than+blood&qid=1765394786&sprefix=thicker+than+blood%2Caps%2C112&sr=8-1)
+ [Weapons of Math Destruction](https://www.amazon.com/Weapons-Math-Destruction-Increases-Inequality/dp/0553418831/ref=sr_1_1?crid=2N1WP8NGZ1XXT&dib=eyJ2IjoiMSJ9._Ib1rdVR1ZEU9t48f_U3Pmins7VDe8N2cKKYOjkNtS6AiOOormoIN6FltOpzqJtx8QU3tHvxjeQPcflsJZl3FVSO76vJeOcPA24Ha2WS61czkcDglr3qHqLE0e6uyb-TKFUIrcABi6DX9jkRpJPZcDfAe73iv8tg5ej_yMYn3fJXcC0WaDpaYLw8a0bA9DViZX5jRNKdmDUGvHsFOWz5fTyWzhWDcwPlfBW9-QROa4k.PCBIMitn9CxJFz4eYVkrMnVmtwOgHNOjY4sp3ufFuiw&dib_tag=se&keywords=weapons+of+math+destruction&qid=1765394803&sprefix=weapons+of+math%2Caps%2C121&sr=8-1)


---

# Miscellaneous resources

+ [Tidyverse style guidelines](https://style.tidyverse.org)
+ Create nicely formatted tables w/ kableExtra
  + [LaTeX / pdf tables](https://haozhu233.github.io/kableExtra/awesome_table_in_pdf.pdf)
  + [HTML tables](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html)

---

# Final thoughts

+ Keep learning

--

+ Keep programming (reproducibly!)

--

+ Keep being thoughtful about data ethics
