---
title: "Lab 01: Hello R!"
subtitle: "SOLUTIONS"
author: "Katie Fitzgerald"
date: "due September 6"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    number_section: no
    highlight: tango
    theme: cosmo
    fig_height: 7
editor_options: 
  chunk_output_type: console
---

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(datasauRus)
```

# Exercise 1

There are `r nrow(datasaurus_dozen)` rows and `r ncol(datasaurus_dozen)` in the `datasaurus_dozen` dataset. The names of the variables are below.

```{r ex-1}
names(datasaurus_dozen)
```

# Exercise 2 

```{r ex-2}
dino_data <- datasaurus_dozen %>%
  filter(dataset == "dino")

ggplot(data = dino_data, mapping = aes(x = x, y = y)) +
  geom_point()

dino_data %>%
  summarize(r = cor(x, y))
```

# Exercise 3

The correlation for the `dino` and `star` datasets are very similar to one another, even though they have very different data when visualized. 

```{r ex-3}
star_data <- datasaurus_dozen %>%
  filter(dataset == "star")

ggplot(data = star_data, mapping = aes(x = x, y = y)) +
  geom_point()

star_data %>%
  summarize(r = cor(x, y))
```

# Exercise 4

The correlation for the `dino` and `circle` datasets are very similar to one another, even though they have very different data when visualized. 

```{r ex-4, fig.height=5, fig.width=5}
circle_data <- datasaurus_dozen %>%
  filter(dataset == "circle")

ggplot(data = circle_data, mapping = aes(x = x, y = y)) +
  geom_point()

circle_data %>%
  summarize(r = cor(x, y))
```

# Exercise 5

```{r ex-5}
ggplot(datasaurus_dozen, aes(x = x, y = y, color = dataset))+
  geom_point()+
  facet_wrap(~ dataset, ncol = 3) +
  theme(legend.position = "none")

datasaurus_dozen %>%
  group_by(dataset) %>%
  summarize(r = cor(x, y)) %>%
  print(13)
```
