---
title: "Data Wrangling III"
author: "YOUR NAME HERE"
date: "today"
format: 
  html:
    self-contained: true
    toc: true
    toc_float: true
    number_section: false
    highlight: "tango"
    theme: "cosmo"
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r setup, message = FALSE}
library(tidyverse)
```

To start, you will work with the following data called `scores`

```{r}
scores <- tribble(
  ~student_id, ~math, ~english,
  "S1",        90,    92,
  "S2",        85,    80,
  "S3",        88,    85,
  "S4",        95,    74
)
scores
```

Note: The `tribble()` function above is helpful for creating small data frames (tibbles) with an easier to read row-by-row layout.

It has three variables (`student_id`, `math`, and `english`) and four rows (one for each student).

# Exercise 1

Before writing any code, answer the following questions:

Suppose you want to reshape the data frame so that there is one row per student per subject. What function would you use to do this?

> `pivot_longer`

If you reshaped the data to have one row per student per subject, how many rows would the resulting data frame have?

> 8

How many columns would the resulting data frame have and what would the column names be?

> 3, `student_id`, `subject`, `score`

Now, write the code to reshape the data frame as described above.

```{r}
scores |> 
  pivot_longer(math:english, names_to = "subject", values_to = "score")
```

# Exercise 2

```{r}
patients <- tribble(
  ~patient_id, ~measurement_time_, ~systolic_bp,
  "P1",        "Morning",          120,
  "P1",        "Noon",             115,
  "P1",        "Evening",          123,
  "P2",        "Morning",          118,
  "P2",        "Evening",          121
)

patients
```

It has three variables (`patient_id`, `measurement_time_`, and `systolic_bp` -- short for systolic blood pressure) and five rows (one per patient per measurement time).

Before writing any code, answer the following questions:

-   Suppose you want to reshape the data frame so that there is one row per patient and measurements at different times of the day are recorded in different columns. What function would you use to do this?

> `pivot_wider`

-   If you reshaped the data to have one row per patient,
    -   how many rows would the resulting data frame have?
    -   how many columns would the resulting data frame have and what would the column names be?

> 2 rows, 3 columns - patient_id, bp_morning, bp_noon, bp_evening

Now, write the code to reshape the data frame as described above. What does the `NA` value mean in the resulting data frame?

```{r}
patients |> 
  pivot_wider(names_from = measurement_time_, 
              values_from = systolic_bp,
              names_prefix = "bp_")
```

> YOUR ANSWER HERE

# grad years

For these next few questions, you will work with the following dataset called `grad_years` as well as the student scores dataset from Exercise 1 (`scores`), and join them with various join functions.

```{r}
#| eval: true
# fmt: skip
grad_years <- tribble(
  ~id,  ~graduation_year,
  "S1", 2023,
  "S3", 2023,
  "S5", 2025,
  "S6", 2024
)

grad_years
```

# Exercise 3

Your friend writes the following code to join the `scores` and `grad_years` data frames and gets an error message:

```{r}
#| eval: true
#| error: true
scores |>
  left_join(grad_years)
```

What does the error message mean and why does it occur?

> no common linking variables

Provide a corrected version below:

```{r}
scores |>
  left_join(grad_years, by = c("student_id" = "id"))
```

How many rows and columns does the resulting data frame have? Explain why.

> 4x4: there are

# Exercise 4

Don't write any code yet: Suppose you join the two data frames, `scores` and `grad_years` with a `right_join()`, in that order. How many rows and columns would the resulting data frame have?

> YOUR ANSWER HERE

Now, write the code to perform the join. If your guess wasn't correct, discuss with your neighbor before proceeding.

```{r}

```

# Exercise 5

Start with the code this time. Join the two data frames, `scores` and `grad_years` with an `inner_join()`, in that order. How many rows and columns does the resulting data frame have? Explain why.

```{r}

```

> YOUR ANSWER HERE

# Exercise 6

Don't write any code yet: Suppose you join the two data frames, `grad_years` and `scores` with an `inner_join()` again, but in the reverse order (`grad_years` first, then `scores`). Would you expect the resulting data frame to have the same number of rows as the previous part or a different number of rows? Explain your reasoning.

> YOUR ANSWER HERE

Now, write the code to perform the join. If your guess wasn't correct, discuss with your neighbor before proceeding.

```{r}

```

# Exercise 7

Don't write any code yet: Suppose you join the two data frames, `scores` and `grad_years` with an `anti_join()`. Which observation(s) would be in the resulting data frame?

> YOUR ANSWER HERE

Now, write the code to perform the join. If your guess wasn't correct, discuss with your before proceeding.

```{r}

```

# Exercise 8

Don't write any code yet: Suppose you join the two data frames, `grad_years` and `scores` with an `anti_join()` again, but in the reverse order (`grad_years` first, then `scores`). Which observation(s) would be in the resulting data frame?

> YOUR ANSWER HERE

Now, write the code to perform the join. If your guess wasn't correct, discuss with your neighbor before proceeding.

```{r}

```

# Exercise 9

Start with the code this time. Join the two data frames, `scores` and `grad_years` with a `full_join()`, in that order. How many rows and columns does the resulting data frame have? Explain why.

```{r}

```

> YOUR ANSWER HERE

# Exercise 10

Your ultimate goal is to construct a dataset where you have all students who are in the `scores` data frame and their subject scores and graduation years. In order to accomplish this, you will need to look up some students' graduation years in an alternate resource, since they're not all in the `grad_years` data frame. Which join function would you use to find out whose graduation years you need to look up in an alternate resource? Write the code to perform the join to confirm your answer.

> YOUR ANSWER HERE

```{r}

```

```{r}
survey_counts <- tribble(
  ~age,    ~opinion,            ~n,
  "18-44", "Very positive",     78,
  "18-44", "Somewhat positive", 176,
  "18-44", "Neutral",           162,
  "18-44", "Somewhat negative", 50,
  "18-44", "Very negative",     36,
  "18-44", "Unsure",            197,
  "45+",   "Very positive",     41,
  "45+",   "Somewhat positive", 121,
  "45+",   "Neutral",           186,
  "45+",   "Somewhat negative", 146,
  "45+",   "Very negative",     97,
  "45+",   "Unsure",            210
)  |>
  mutate(
    opinion = factor(opinion, levels = c("Very positive",
                                              "Somewhat positive",
                                              "Neutral",
                                              "Somewhat negative",
                                              "Very negative",
                                              "Unsure"))
  )

survey_counts |>
  pivot_wider(
    names_from = "age",
    values_from = n
  )
```

## Exercise 3

Marginal proportions of age: Calculate the proportions of individuals who are 18-44 year olds and 45+ year-olds in this sample.

```{r}
survey_counts |> 
  group_by(age) |> 
  summarize(n = sum(n)) |> 
  mutate(prop = n/sum(n))
```


c.  Marginal proportions of opinion: Calculate the proportions of individuals who are Very positive, Somewhat positive, Neutral, Somewhat negative, Very negative, and Unsure.

```{r}
survey_counts |> 
  group_by(opinion) |> 
  summarize(n = sum(n)) |> 
  mutate(prop = n/sum(n))
```


d.  Conditional proportions of opinion based on age: Calculate the proportions of individuals who are Very positive, Somewhat positive, Neutral, Somewhat negative, Very negative, and Unsure

    -   among those who are 18-44 years old and
    -   among those who are 45+ years old.
    
```{r}
survey_counts |> 
  group_by(age) |> 
  mutate(prop = n/sum(n))
```

Adapt your code above to display your proportions as a 6x3 data frame where the first column lists the opinion category, and the 2nd and 3rd columns give the proportions for the `18-44` and `45+` age categories, respectively.

```{r}
survey_counts |> 
  group_by(age) |> 
  mutate(prop = n/sum(n)) |> 
  pivot_wider(id_cols = opinion, names_from = age, 
              values_from = prop)
```




:
