---
title: "Working with strings"
author: "YOUR NAME HERE"
date: "today"
format: 
  html:
    self-contained: true
    toc: true
    toc_float: true
    number_section: false
    highlight: "tango"
    theme: "cosmo"
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r setup, message = FALSE}
library(tidyverse)
```

# State names

R has built-in vectors with the names of all 50 states in the U.S.

```{r}
state.name
```

## Exercise 1

Run each line below and add comments to the code to indicate what each line is doing.

```{r}
str_which(state.name,"V")
str_subset(state.name,"V")
str_subset(str_to_upper(state.name),"V")
str_subset(state.name,"[Vv]")
```

## Exercise 2

Plot n over time by name for girls whose name starts with “Elizab” but isn’t “Elizabeth”

-   read in data
-   need to put this inside filter - str_detect returns logical, works well with filter
-   str_detect(name, "\^Elizab") \^ starts with
-   str_sub(name, 1, 6) == "Elizab"
-   ask ChatGPT to get to: str_detect(name, "\^Elizab(?!eth\$)")

```{r}
babynames <- read_csv("./data/Babynames.csv")

test <- babynames |> 
  filter(str_detect(name, "^Elizab")) |> 
  filter(name != "Elizabeth")

test2 <- babynames |> 
  filter(str_sub(name, 1, 6) == "Elizab") |> 
  filter(name != "Elizabeth")

test3  <- babynames |> 
  filter(str_detect(name, "^Elizab(?!eth$)"))

ggplot(test, aes(x = year, y = n, color = name)) + 
  geom_line()

test |> distinct(name)

```

## Exercise 3

How many unique names are there that contain the string “liz”

```{r}
liz <- babynames |> 
  filter(str_detect(name, "liz")) |> 
  distinct(name)
```

## Exercise 4

Explore variations of your own name

```{r}
katie <- babynames |> 
  filter(str_detect(name, "Katie")) |> 
  filter(sex == "F")

ggplot(katie, aes(x = year, y = n, color = name)) +
  geom_line()

kaitlyn <- babynames |> 
  filter(str_detect(name, "Kaitlyn")) |> 
  filter(sex == "F")

ggplot(kaitlyn, aes(x = year, y = n, color = name)) +
  geom_line()

broad <- babynames |> 
  filter(sex == "F") |> 
  filter(str_detect(name, "^(Kat(?!h)|Kati|Kait|Kate)"))

broad |> 
  distinct(name) |> View()
```

## Exercise 5

Title length Movies with question marks Movies with colons Movies with numbers in the titles

```{r}
movies <- movies |> 
  mutate(title_length = str_length(title))

ggplot(movies |> filter(votes > 100), aes(x = title_length, y = rating)) +
  #geom_jitter(alpha = 0.2)
  geom_hex()

movies |> 
  filter(str_detect(title, "\\?")) |> View()
```
