---
title: "Working with strings"
author: "YOUR NAME HERE"
date: "today"
format: 
  html:
    self-contained: true
    toc: true
    toc_float: true
    number_section: false
    highlight: "tango"
    theme: "cosmo"
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r setup, message = FALSE}
library(tidyverse)
```

# State names

R has built-in vectors with the names of all 50 states in the U.S.

```{r}
state.name
```

## Exercise 1

Run each line below and add comments to the code to indicate what each line is doing.

```{r}
str_which(state.name,"V")
str_subset(state.name,"V")
str_subset(str_to_upper(state.name),"V")
str_subset(state.name,"[Vv]")
```

## Exercise 2

Plot n over time by name for girls whose name starts with “Elizab” but isn’t “Elizabeth”



## Exercise 3

How many unique names are there that contain the string “liz”? 



## Exercise 4

Explore variations of your own name


## Exercise 5

What is the most common letter of first names? Has this changed over time? 

## Exercise 6

Let's explore the `movies` dataset from the `ggplot2movies` package. 

+ Do movies with longer titles tend to have higher or lower ratings? 
+ Movies with question marks in the title
+ Movies with colons in the title
+ etc etc


